---
title: "Bladder Cancer in the United States: National Trends and State-Level Patterns from Global Burden of Disease (GBD), 1980-2021"
subtitle: "Burden of Bladder Cancer in the United States"
author: 
  - name: Alireza Sadeghi
    corresponding: true
    orcid: 0000-0002-7950-3270
    email: alireza.sadeghi.md@gmail.com
    degrees: MD
    affiliations: 
      - ref: shiraz
    roles: [conceptualization, data curation, formal analysis, investigation, methodology, software, visualization, writing – review & editing, writing – original draft]
  - name: Fatemeh Nouri
    orcid: 0000-0002-4878-7848
    corresponding: false
    email: nouri.fatmh@gmail.com
    degrees: MD
    roles: [writing – review & editing, writing – original draft, investigation]
    affiliations: 
      - ref: shiraz
  - name: Niloofar Dehdari Ebrahimi
    orcid: 0000-0002-9866-8361
    corresponding: false
    email: niloofar.dehdari6021@gmail.com
    degrees: MD
    roles: [writing – original draft, writing – review & editing, data curation, investigation]
    affiliations: 
      - ref: shiraz
  - name: Ehsan Taherifard
    orcid: 0000-0002-8438-4990
    corresponding: false
    email: etaherim123@gmail.com
    degrees: MD
    roles: [writing – review & editing, resources, data curation, investigation]
    affiliations: 
      - ref: shiraz
  - name: Wassim Kassouf
    corresponding: true
    orcid: 0000-0002-8350-0531
    email: Wassim.Kassouf.med@ssss.gouv.qc.ca
    degrees: MD
    roles: [writing – review & editing, supervision, project administration, investigation]
    affiliations: 
      - ref: mcgill
affiliations: 
  - id: shiraz
    name: Student Research Committee, Shiraz University of Medical Sciences, Shiraz, Iran
  - id: mcgill
    name: Division of Urology, McGill University Health Center, Montreal, QC
keywords: 
  - bladder cancer
  - global burden of disease
  - United States of America
  - incidence
  - mortality
  - prevalence
number-sections: true
bibliography: bladder.bib
csl: vancouver.csl
# format:
#   docx:
#     reference-doc: reference.docx
# filters: 
editor_options: 
  chunk_output_type: console
editor:
  markdown:
    canonical: false
    wrap: 100
---

```{css}
#| echo: false
.body-justified p {
  text-align: justify
}
```


```{r setup}
#| include: false
#| echo: false
#| message: false 
#| warning: false

# Setting up the quarto notebook

options(
  tidyverse.quiet = TRUE,
  dplyr.summarise.inform = TRUE
)

library(tidyverse)
library(scales)
library(glue)
library(geofacet)
library(ggnewscale)
library(future)
library(furrr)
library(gganimate)
library(paletteer)
library(flextable)

bladder_data <- 
  paste0("data/", list.files("data", "rds")) %>%
  map_df(read_rds, .progress = T) %>% 
  bind_rows()


stateids <- 
  bladder_data %>% 
  filter(location_level == "United States of America") %>% 
  distinct(location_id) %>%
  as_vector()

source("scripts/functions.r")


```


```{r}
#| output: false
#| results: false

# Creating the geofacet plot

geofac_data <- 
  bladder_data %>% 
  filter(location_id %in% c(stateids, 102)) %>% 
  group_by(measure_name, cause_name, metric_name) %>% 
  nest() %>% 
  ungroup()


pmap(geofac_data, geofacet_state_plot, .progress = T)

```



```{r}
#| output: false
#| results: false

# Creating the plot for comparing US performance compared to competing countries


## Line plot


us_performance_data <- 
  bladder_data %>% 
  filter(
    location_id %in% c(4743, 64, 102, 44635),
    sex_id == 1 & cause_id %in% c(468, 438) |
      sex_id == 3 & cause_id %in% c(474, 471)
  ) %>% 
  mutate(
    sex_id = NULL,
    sex_name = NULL,
    stateabb = NULL,
    location_level = NULL,
    country_group = if_else(
      location_id == 102, 
      "United States of America",
      "European Union/High-income/High SDI"
    )
  ) %>% 
  group_by(cause_name, metric_name) %>% 
  nest %>% 
  ungroup() %>% 
  mutate(
    data = 
      map(data, \(x) {
        
        if (27 %in% x$age_id) {
          
          data <- 
            x %>% 
            filter(age_id == 27)
          
          attr(data, "ageattr") <- "Age-standardized"
          
        } else {
          
          data <- 
            x %>% 
            filter(age_id == 22)
          
          attr(data, "ageattr") <- "All Ages"
          
        }
        
        return(data)
      }, .progress = T)
  )



pmap(us_performance_data, us_performance_plt, .progress = T)

```


```{r}
#| output: false
#| results: false


## Bump plot


states <- 
  bladder_data %>% 
  filter(!is.na(stateabb)) %>% 
  distinct(location_id, location_name, stateabb)

bumpyears <- 
  c(1990, 2000, 2010, 2021)

bump_data <- 
  bladder_data %>% 
  group_by(cause_name, metric_name) %>% 
  nest %>% 
  ungroup %>% 
  mutate(
    data = map(
      data, 
      \(x) {
        
        cdata <- 
          x %>% 
          select(-c(upper, lower)) %>% 
          filter(
            year %in% bumpyears,
            location_id %in% c(4743, 64, 102, 44635, stateids),
            sex_id == 1 & cause_id %in% c(468, 438) |
              sex_id == 3 & cause_id %in% c(474, 471),
            age_id %in% c(22, 27)
          )
        
        agedata <- 
          cdata %>% 
          group_by(location_id, .add = T) %>% 
          count(age_id) %>% 
          filter(age_id %in% c(22, 27)) %>%
          arrange(age_id, n) %>%
          group_by(age_id) %>% 
          summarise(sum = sum(n))
        
        if (27 %in% agedata$age_id) {
          
          cdata <- 
            cdata %>% 
            filter(age_id == 27)
          
          ageattr <- "Age-standardized"
          
        } else {
          
          cdata <- 
            cdata %>% 
            filter(age_id == 22)
          
          ageattr <- "All Ages"
          
        }
        
        cdata <- 
          cdata %>% 
          group_by(measure_name, year) %>% 
          arrange(val) %>% 
          mutate(rank = row_number()) %>% 
          arrange(measure_id, year, rank) %>% 
          mutate(order = row_number()) %>% 
          ungroup() %>% 
          mutate(
            color = if_else(
              location_id == 102 |
                (
                  year == max(.$year) & 
                    !is.na(stateabb) & 
                    !(order %>% between(6, 50))
                ), 
              "yes", "no"),
            stateabb = if_else(
              is.na(stateabb),
              location_name, stateabb
            ), 
            stateabb = if_else(
              stateabb == "United States of America",
              "USA", stateabb
            ),
            stateabb = if_else(
              stateabb == "European Union",
              "EU", stateabb
            )
          ) %>% 
          group_by(location_name, measure_name)
        # mutate(n = n()) %>% 
        # filter(n != 1) %>% 
        # mutate(n = NULL)
        
        attr(cdata, "ageattr") <- ageattr
        
        return(cdata)
        
      }, .progress = TRUE
    )
  )

pmap(bump_data, bumpplot, .progress = T)

```



```{r}
#| output: false
#| results: false

# Population pyramid plots

## Data preparation


pyramid_data <- 
  bladder_data %>% 
  filter(
    age_id %in% c(1, 6:21),
    sex_id != 3
  ) %>% 
  select(-location_level) %>% 
  arrange(age_id) %>% 
  mutate(age_name = fct_inorder(age_name)) %>% 
  group_by(
    # cause_name, 
    measure_name, 
    # metric_name,
    location_name,
  ) %>% 
  nest() %>% 
  filter(
    !(location_name %in% 
        c(
          "Middle SDI",
          "G20", 
          "Low SDI",
          "Global", 
          "High-middle SDI",
          "Low-middle SDI"
          )
      )
  )

```


```{r}
#| output: false
#| results: false


## Population pyramids for 2021

plan(multisession, workers = 6)

future_pmap(
  pyramid_data, 
  \(data, measure_name, location_name) {
    
    pyramid_filename <- 
      paste0(
        c(
          "fig_C",
          location_name, 
          "Bladder cancer",
          measure_name, 
          "Rate.svg"
          ), 
        collapse = "_"
        )
    
    if (
      file.exists(
        paste0(
          "output/figures/Population pyramid - 2021/", 
          pyramid_filename
          )
        )
      ) {return()}
    
    pyramid_plot(
      data = data,
      measure_name = measure_name
      ) %>% 
      ggsave(
        plot = .,
        device = "svg",
        path = "output/figures/Population pyramid - 2021/",
        filename = pyramid_filename,
        width = 25,
        height = 25,
        units = "cm",
        dpi = 600
        )
    },
  .progress = TRUE
  )

plan(sequential)

```


```{r}
#| output: false
#| results: false

## Population pyramid - animated 1980-2021


years <- 
  bladder_data %>% 
  distinct(year) %>%
  pull


plan(multisession, workers = 6)


future_pmap(pyramid_data,
  
  \(
    data,
    measure_name,
    location_name
  ) {
    
    file_name <-
      paste0(
        c(
          location_name,
          "Bladder cancer",
          measure_name,
          "Rate"
        ),
        collapse = "_"
      ) %>% 
      paste0(
        "output/figures/Population pyramid - animated/fig_E_",
        ., 
        ".gif"
        )
    
    if (file.exists(file_name)) {return(NA)}
    
    library(gganimate)
    
    plot <- 
      pyramid_plot(
        data = data, 
        measure_name = measure_name, 
        year = years
      )
    
    
    iserror <- 
      tryCatch(
        {
          ggplot_build(plot)
          FALSE
        },
        error = function(e) {
          TRUE
        }
      )
    
    if (iserror) {
      beepr::beep()
    }
    
    plot <-
      plot +
      transition_time(year) +
      labs(title = 'Year: {as.integer(frame_time)}') 
    
    anim_save(
      filename = file_name,
      animation = plot,  
      device = "png",
      width = 25,
      height = 25, 
      units = "cm", 
      res = 300
    ) 
    
    
  }, 
  .progress = TRUE
)


plan(sequential)

```


```{r}
#| output: false
#| results: false


# APC Plots


apc_data <- 
  bladder_data %>% 
  filter(
    year %in% c(1990, 2000, 2010, 2021),
    age_id %in% c(1, 6:21),
    location_id %in% c(stateids, 102),
    sex_id != 3,
    metric_id == 3
    ) %>% 
  arrange(age_id, year) %>% 
  mutate(
    age_name = fct_inorder(age_name)
    ) %>% 
  group_by(location_name, measure_name) %>% 
  nest

plan(multisession, workers = 4)

future_map(list(apc_plot1, apc_plot2), \(x) pmap(apc_data, x), .progress = TRUE)

plan(sequential)



```

```{r}
#| output: false
#| results: false


# Table

file_name <-
  paste0(
    "output/tables/Percent change since 1980 in ",
    "Bladder cancer",
    " measures.docx"
  )

change_data <-
    bladder_data %>%
    filter(
      year %in% c(2021, 1990),
      metric_id == 3,
      age_id == 27,
      sex_id != 3,
      !(location_id %in% c(1, 44639, 44634, 44586, 44636, 44637))
    ) %>%
    group_by(
      measure_name, location_name, sex_name, cause_name
    ) %>%
    select(1:6, 9, 10, 13:16, 18) %>%
    # replace true value by deviations
    mutate(upper = upper - val, lower = val - lower) %>%
    pivot_wider(
      id_cols = c(
        measure_id,
        measure_name,
        location_id,
        location_name,
        sex_id,
        sex_name,
        cause_id,
        cause_name,
        stateabb
      ),
      names_from = year,
      values_from = c(val, upper, lower),
      names_glue = "{.value}_{year}"
    ) %>%
    mutate(
      # Propagation of uncertainities
      # references in
      # https://chatgpt.com/share/68317e19-1f4c-800e-a8db-ec53c8621521
      diffval = ((val_2021 - val_1990) / val_1990),
      diffupper = diffval + ((sqrt((upper_2021^2) + (lower_1990^2))) / val_1990),
      difflower = diffval - ((sqrt((lower_2021^2) + (upper_1990^2))) / val_1990)
    ) %>%
    select(-c(val_1990:lower_2021)) %>%
    ungroup() %>%
    mutate(
      color = case_when(
        diffupper > 0 & difflower > 0 ~ "positive",
        diffupper < 0 & difflower < 0 ~ "negative",
        sign(diffupper) != sign(difflower) ~ "neutral",
      )
    ) %>%
    group_by(cause_name, measure_name) %>%
    mutate(
      upperlim = max(diffupper),
      lowerlim = min(difflower)
    ) %>%
    ungroup() 

if (!file.exists(file_name)) {
  

  table_ggplot_data <-
    change_data %>%
    mutate(
      plot = pmap(
        list(diffval, diffupper, difflower, upperlim, lowerlim, color),
        ggmaker,
        .progress = TRUE
      )
    ) %>% 
    arrange(desc(is.na(stateabb)), location_name) %>%
    mutate(
      location_name = if_else(is.na(stateabb), location_name, stateabb)
    ) %>%
    select(
      -c(diffval:lowerlim),
      -contains("_id"),
      -stateabb
    ) %>%
    rename_with(\(x) str_remove(x, "_name")) %>%
    pivot_wider(
      names_sep = "_",
      id_cols = c(location, cause),
      names_from = c(measure, sex),
      values_from = plot
    ) %>%
    group_by(cause) %>%
    nest() %>%
    mutate(
      data = if_else(
        cause %in% c("Prostate cancer", "Testicular cancer"),
        map(data, \(x) {
          x %>%
            select(-contains("Female"))
        }),
        data
      )
    )


  table_ggplot_data %>%
    mutate(
      new = map2(
        data,
        cause,
        \(x, y) fpmaker(x, y),
        .progress = TRUE
      )
    )
  
}



```

**Authors contact**

```{r}
#| label: author-contacts
#| echo: false
#| message: false 
#| warning: false
#| results: asis


authors <- rmarkdown::metadata$author

author_data <- 
  authors |> 
  map(\(a) {
    tibble(
      name         = a$name,
      corresponding = a$corresponding %||% NA,
      email        = a$email %||% NA,
      orcid        = a$orcid %||% NA,
      degrees      = a$degrees %||% NA,
      roles        = list(a$roles),
      affiliations = list(map_chr(a$affiliations, "ref"))
    )
  }) |> 
  list_rbind() |>       
  unnest(c(roles, affiliations))


contacts_data <- 
  author_data %>% 
  distinct(name, email, orcid)

cat(
  glue_data(
  contacts_data, 
  "{name}\n\n   + Email <{email}>\n\n  +  ORCID [{orcid}](https://orcid.org/{orcid})\n\n\n", 
  .na = ""
  )
)

```

{{< pagebreak >}}

```{r}
#| label: source-data
#| echo: false
#| message: false
#| warning: false
#| results: hide



bladder_data <-
  bladder_data %>%
  mutate(
    readable_name = case_when(
      measure_name == "Deaths" ~ "mortality",
      measure_name == "Prevalence" ~ "prevalence",
      measure_name == "Incidence" ~ "incidence",
      .default = measure_name
    )
  )

diff_abstract <-
  bladder_data %>%
  filter(
    sex_name == "Both",
    year %in% c(1990, 2021),
    location_name == "United States of America",
    age_name == "Age-standardized"
    ) %>%
  group_by(measure_name) %>%
  select(measure_name, readable_name, year, val) %>%
  pivot_wider(names_from = year, values_from = val) %>%
  rowwise() %>%
  mutate(
    diff = `2021` - `1990`,
    than = ifelse(diff > 0, "lower", "higher"),
    diff = paste(
      diff %>%
        abs %>%
        scales::number(accuracy = 0.01),
      than
    )
    )

glue::glue_data(
  diff_abstract,
  "In 2021, United States recorded age-standardized rates of {scales::number(2021, accuracy = 0.01)} for {readable_name}, which were {diff} than the 1990 records, respectively."
)


items <-
  diff_abstract %>%
  glue::glue_data("{scales::number(`2021`, accuracy = 0.01)} for {readable_name} ({diff})")

# Create the final sentence
final_sentence <-
  glue::glue(
    "In 2021, United States recorded age-standardized rates of {knitr::combine_words(items, and = 'and ', sep = ', ')} compared to 1990 records."
)

```


```{r}
#| output: false
#| results: false


# State abbreviations and full names for captions

state_abbs <- 
  states %>%
  arrange(location_name) %>% 
  glue_data("{location_name}: {stateabb}; ") %>% 
  paste0(collapse = "")

country_abbs <- 
  "European Union: EU; SDI: Socio-Demographic Index"

measure_abbs <- 
  "DALYs: Disability-Adjusted Life Year; YLDs: Years Lived with Disability; YLLs: Years of Life Lost"

```

```{r}
#| output: false
#| results: false


# Supplementary file links

base_url <- 
  "https://github.com/alireza5969/bladder-cancer-gbd"

figures_base_url <- 
  paste0(
  base_url,
  "output/figures/",
  list.files("output/figures")
  )

table_base_url <- 
  paste0(
  base_url,
  "output/tables/"
  )

```

::::: {.body-justified}

# Abstract

**Background:** United States is the country with most expenditure in health care. Yet, the
statistics do not correlate with the amount of money being spent. Investigating national- and
state-level historical trends and comparing them with other countries provides insights for policy
makers. 
**Aim:** We aim to explore the chronological trends in the rate of bladder cancer burden,
including prevalence, incidence, mortality, Disability-Adjusted Life Year (DALY), Years Lived with
Disability (YLD), and Years of Life Lost (YLL). State-level disparities and long-term effects of
health policies were explored. 
**Methodology:** Data from Global Burden of Disease (GBD) was
retrieved. Age- and gender-classified historical trends visualized for each state as well as the
country. Trends were compared with other competing health systems, including European Union,
countries with high Socio-Demographic Index (SDI), and high income.
**Results:** `r final_sentence`
**Conclusion:** To reduce the significant
and disparate burden of BC in the United States, targeted public health
initiatives and region-specific policies are essential. 
These must address modifiable risk factors, improve diagnostic
equity, and expand access to early detection and care.

**Keywords:** *`r rmarkdown::metadata$keywords %>% paste0(collapse = "; ")`*

{{< pagebreak >}}

# Introduction

Bladder cancer (BC) is the sixth most common cancer in the United States [@RN1]. Among American men, it
is the fourth most common cancer and the seventh leading cause of cancer-related mortality [@RN2].
According to the recent Surveillance, Epidemiology, and End Results (SEER) Program estimates,
approximately 84,870 new cases and 17,420 deaths are projected in 2025, accounting for 4.2% of all
new cancer diagnoses and 2.8% of cancer-related mortality [@RN1]. Similar trends are observed
outside the United States, specifically, a rising global burden despite declining age-standardized
rates, highlighting that BC is a significant public health concern [@RN3].

In addition, BC imposes a high economic burden as management typically involves lifelong
surveillance and repeated interventions, making it the costliest cancer to treat per patient over a
lifetime [@RN4; @RN5; @RN6]. Beyond direct healthcare costs, the disease significantly affects
patients' quality of life as well as the emotional and practical well-being of their caregivers and
families [@RN7].

While cigarette smoking has historically been the primary risk factor for BC [@RN8], emerging
evidence points to a widening array of carcinogenic exposures contributing to its global incidence.
These include, but are not limited to, arsenic-contaminated drinking water, occupational and
cosmetic chemicals (such as those in hair dyes), dietary factors, and pervasive environmental
pollutants [@RN9]. Despite observing a decrease in the prevalence of cigarette smoking since 1998
[@RN10] and the implementation of stricter government regulations on the use of aryl amines, there
has not been a corresponding reduction in the incidence of BC [@RN11]. This stagnation suggests that
other environmental and occupational exposures may be driving disease patterns, particularly in
regions with inadequate regulatory oversight. One example is the elevated cancer incidence rates
observed in certain New England states, which have been partly attributed to environmental
carcinogens such as arsenic contamination in drinking water from private household wells [@RN12;
@RN13]. The complex interplay between these exposures and persistent regional and socioeconomic
disparities underscores a pressing public health concern; one that remains inadequately
characterized and urgently demands targeted surveillance, research, and policy intervention.

Previous studies have demonstrated substantial regional disparities in BC incidence and outcomes
within the United States [@RN14; @RN15; @RN16; @RN17]. For example, Maine reports an
age-standardized incidence rate (ASIR) of 26.7 between 2017 and 2021, nearly double that of Hawaii,
which stands at 13.4 [@RN18]. Geographic variations likely stem from differences in environmental
exposures, socioeconomic status, access to care, and healthcare infrastructure. For instance, BC
incidence is highest in northeastern states [@RN16], while mortality tends to be greater in Southern
states [@RN19] where there is higher proportion of older adults, African Americans, and rural
populations, groups that often face delays in diagnosis and limited access to quality treatment
[@RN20].
<!-- Additionally, while men are more frequently diagnosed, women often experience worse -->
<!-- outcomes, including higher stage at diagnosis and lower survival rates [@RN21]. -->

While most BC research to date has focused on national trends or single-institution cohorts, recent,
age-standardized comparisons at the state level are scarce. Likewise, although the Global Burden of
Disease (GBD) project produces Disability-Adjusted Life Year (DALYs), Years Lived with Disability
(YLDs), and Years of Life Lost (YLLs) for BC, these metrics have seen limited uptake in
disease-specific epidemiological studies, leaving non-fatal burden and health-system impact
underexplored. National analyses report that men have roughly three to four times the incidence of
BC [@RN22] yet women present with more advanced disease and poorer survival [@RN23]; however, these
sex-specific outcome gaps have not been systematically mapped across the states.

To address these gaps, we utilize data from the GBD 2021 study, which provides consistent estimates
of key burden metrics for all states. This mostly-descriptive analysis offers a comprehensive
insight of BC burden across geographic, temporal, and age group dimensions.

<!-- By incorporating [age--period--cohort]{.mark} modeling and evaluating -->

<!-- sex-specific and socio-demographic patterns, our study aims to generate actionable insights for -->

<!-- state-level public health planning and prioritization efforts. -->

# Methodology

## Study Design and Data Sources

This population-based study used data from the GBD 2021 database
<!-- , compiled by the Institute for Health Metrics and Evaluation (IHME)  --> . GBD offers
comprehensive, modeled estimates for all states, enabling standardized comparisons across time,
geography, and socioeconomic strata, making it well-suited for analyzing disparities in BC burden on
a national and global scale. The study analyses BC (ICD-10: C67) burden across all the states of the
US from 1980 to 2021, assessing incidence, prevalence, mortality, YLLs, YLDs, and DALYs.
Age-standardized rates (/100,000) were reported unless data were explicitly stated otherwise. 
The methodology of the GBD was consistent with previous publications and the details provided in the
Appendix. This study complies with the Guidelines for Accurate and Transparent Health Estimates
Reporting (GATHER) statement.

## Definitions

DALYs

:   A measure of the gap in healthy years of life lived by a population as compared with a normative
    standard. More formally, DALYs are a time based measure which adds together years of life lost
    due to premature mortality with the equivalent number of years of life lived with disability or
    illness.

YLLs

:   The component of the DALYs that measures years of life lost due to premature mortality.

YLDs

:   The component of the DALYs that measures lost years of healthy life through living in health
    states of less than full health.

Socio-Demographic Index (SDI)

:   A composite average of the rankings of the incomes per capita, average educational attainment, and fertility rates of all areas in the GBD study.
    
Competing Countries
:   This study defines the groups of countries that are comparable with the United States 
    in the sense of development and income. These country groups include 
    European Union, high-SDI countries, and high-income countries.

## Statistical Analysis

We reported age-standardized rates for each burden metric using the direct method with the GBD 2021
world standard population. Age-standardized rates were not applicable for *age- and sex-specific
rate curves*. Temporal trends were calculated since 2018 until 2021. All rates were expressed per
100,000 population. 
Uncertainty intervals, used by the GBD study’s Bayesian framework, were defined
as the 2.5th and 97.5th percentiles of 500 posterior draws from the framework and account for
multiple sources of uncertainty [@naghavi2024].

We used R version 4.5.1 (2025-06-13 ucrt) for calculations. In addition, 28 R packages were employed
to facilitate data curation, visualization, analysis, and reproducibility [@allaire2024b; @allaire2024c; @atkins2025b; @attali2021a; @bååth2024a;
@campitelli2025a; @chang2025a; @gohel2025a; @hafen2025; @hester2024; @hughes2022; @hvitfeldt2021;
@müller2025; @pedersen2025; @richardson2025; @RJ-2021-048; @sievert2025; @sjoberg2020; @vaughan2022;
@wickham2016; @wickham2019; @wickham2023; @wickham2025; @wilke2022; @xie2014a; @xie2015; @xie2018;
@xie2020; @xie2025]. 
The full methodology, including data, code, and notebooks are
detailed in the study's [GitHub repository](https://github.com/alireza5969/bladder-cancer-gbd).

# Results

## Bladder Cancer Burden: United States Compared to Competing Countris

```{r}

incidence_death_us_2021 <-
  bladder_data %>%
  filter(
    location_name == "United States of America", 
    age_name == "Age-standardized",
    year == 2021,
    measure_name %in% c('Incidence', "Deaths"),
    sex_name != "Both"
    ) %>% 
  select(measure_name, sex_name, val, upper, lower) 


incidence_death_us_2021_txt <- 
  glue_data(
    incidence_death_us_2021,
    "{number(val, .1)} [{number(lower, .1)}, {number(upper)}]"
  )

incidence_death_compete_2021 <- 
  bladder_data %>%
  filter(
    is.na(stateabb), 
    age_name == "Age-standardized",
    year == 2021,
    measure_name %in% c('Incidence', "Deaths"),
    sex_name == "Both"
    ) %>% 
  select(measure_name, location_name, val, upper, lower)


incidence_death_compete_2021_txt <- 
  glue_data(
    incidence_death_compete_2021,
    "{number(val, .1)} [{number(lower, .1)}, {number(upper)}]"
  )

```

In 2021, the incidence rates were `r incidence_death_us_2021_txt[1]`
and `r incidence_death_us_2021_txt[2]`, resulting in 
a mortality rate of `r incidence_death_us_2021_txt[3]` and `r incidence_death_us_2021_txt[4]`, respectively.
In the same year, European Union, high-income, and high-SDI countries 
reported mortality rates of `r incidence_death_compete_2021_txt[5]`, 
`r incidence_death_compete_2021_txt[7]`, 
and 
`r incidence_death_compete_2021_txt[6]`, respectively
in both genders.



```{r}


max_vals_us <- 
  bladder_data %>%
    filter(
        measure_name %in% c('Incidence', "Prevalence"),
        location_name == "United States of America",
        age_name == "Age-standardized", 
        sex_name == "Both"
    ) %>% 
    group_by(measure_name) %>% 
    slice_max(val) %>% 
    select(val, upper, lower, year)

max_vals_us_txt <- 
  glue_data(
    max_vals_us,
    "{number(val, .1)} [{number(lower, .1)}, {number(upper)}]"
    )

```

With the highest age-standardized 
rates of `r max_vals_us_txt[1]` and `r max_vals_us_txt[2]` in 2011 for 
incidence and prevalence, respectively, United States has
almost consistently had  
higher incidence and prevalence of BC compared to 
competing countries between 1980 and 2021. 
In contrast, DALYs, YLLs, and mortality rates 
have been lower than competing countries 2000s.
The clear declining trend observable in competing countries 
is absent in the trend of United States. 
While these rates have been historically low in United States, 
competing countries have managed to 
approach to the statistics of United States (@fig-performance).

![Age-standardized bladder cancer burden rates between 1980 and 2021 in the United States compared to high-income and high socio-demographic index countries and European Union.](output/figures/US performance plot/fig_B_Bladder cancer_Rate.svg){#fig-performance}


```
Abbreviations: `r measure_abbs`
```


```{r}


country_change_data <- 
  change_data %>% 
  filter(
    is.na(stateabb),
    ) %>% 
  mutate(
    sex_name = if_else(sex_id == 2, "women", "men"),
    across(diffval:difflower, \(x) return(abs(x) * 100))
    )

green_change_data <- 
  country_change_data %>% 
    filter(
        color == "negative",
        measure_name %in% c(
          "DALYs", 
          "YLLs",
          "Deaths"
          # "Incidence"
          )
    ) %>% 
    pivot_wider(
      values_from = diffval:difflower,
      names_from = sex_name,
      id_cols = c(measure_name, location_name)) %>%
  arrange(measure_name, location_name)

red_change_data <- 
  country_change_data %>% 
  filter(color == "positive")

green_change_txt <- 
  glue_data(
    green_change_data,
    "men: {number(diffval_men, .1)}% [{number(difflower_men, .1)}%, {number(diffupper_men, .1)}%]; women: {number(diffval_women, .1)}% [{number(difflower_women, .1)}%, {number(diffupper_women, .1)}%]", 
    .na = ""
    )

incidence_green_change_data <- 
  country_change_data %>% 
    filter(
        color %in% c("negative", "neutral"),
        measure_name == "Incidence",
        sex_name == "men"
        )

incidence_green_change_txt <- 
  glue_data(
    incidence_green_change_data,
    "{number(diffval, .1)}% [{number(difflower, .1)}%, {number(diffupper, .1)}%]", 
    .na = ""
    )

red_change_txt <- 
  glue_data(
    red_change_data,
    "{number(diffval, .1)}% [{number(difflower, .1)}%, {number(diffupper, .1)}%]"
    )

```

A more abstract but quantified version of data
in @fig-performance could be seen in
@tbl-percentchange which shows the percentage of age-standardized
rates changed between 1980 and 1990 for all measures and across all states.
Similar to @fig-performance, the table visualizes decreased rates in
DALYs (`r green_change_txt[3]`, `r green_change_txt[2]`, 
`r green_change_txt[1]`, and `r green_change_txt[4]`)
and 
YLLs (`r green_change_txt[10]`, `r green_change_txt[9]`, 
`r green_change_txt[8]`, and `r green_change_txt[11]`)
across high-income and high-SDI countries, European Union,
and United States, respectively.
On the other hand, while high-income and high-SDI countries, and 
European Union managed to significantly 
decrease their statistics in mortality 
(`r green_change_txt[7]`, `r green_change_txt[6]`, and 
`r green_change_txt[5]`,
respectively) and 
male incidence
(`r incidence_green_change_txt[3]`, `r incidence_green_change_txt[4]`,
`r incidence_green_change_txt[1]`),
United States shows insignificant changes
(`r incidence_green_change_txt[2]`).

![Percentage change of age-standardized rates between 1980 and 2021 across the United States, compared to other competing countries. Bars and error lines indicate the average percentage change and its 95% confidence interval, respectively. Green, yellow, and red indicate significant decrease, insignificant change, and significant increase in statistics, respectively.](output/tables/Percent change since 1980 in Bladder cancer measures.svg){#tbl-percentchange}

```
Abbreviations: `r paste0(c(measure_abbs, country_abbs, state_abbs), collapse = "; ")`
```


As demonstrated in @fig-pyramid, in 2021, 
all burden measures were 
higher in men in all age groups. 
Highest rates of each burden measure among the 
age groups
are as follows:

```{r}
#| results: asis


pyramid_results_data <- 
  pyramid_data %>% 
  filter(location_name == "United States of America") %>% 
  mutate(
    data = map(data, \(x) {
      x %>% 
        group_by(sex_name) %>% 
        slice_max(val) %>% 
        ungroup() %>% 
        select(sex_name, age_name, val:lower)
    })
  ) %>% 
  unnest(data) %>% 
  pivot_wider(
    id_cols = c(measure_name, age_name),
    names_from = sex_name,
    values_from = val:lower
    ) 
# %>% mutate(tag = paste0("{#fig-pyramid-", tolower(measure_name), "}"))


pyramid_results_txt <- 
  pyramid_results_data %>% 
  glue_data(
    "* {measure_name}: {number(val_Male, .1, big.mark = ',')} [{number(lower_Male, .1, big.mark = ',')}, {number(upper_Male, .1, big.mark = ',')}] in men and {number(val_Female, .1, big.mark = ',')} [{number(lower_Female, .1, big.mark = ',')}, {number(upper_Female, .1, big.mark = ',')}] in women in {age_name} age group.\n\n\n"
  )

cat(pyramid_results_txt)

```


::: {#fig-pyramid layout-ncol=3}

![Prevalence](output/figures/Population pyramid - 2021/fig_C_United States of America_Bladder cancer_Prevalence_Rate.svg){#fig-pyramid-prevalence}

![Incidence](output/figures/Population pyramid - 2021/fig_C_United States of America_Bladder cancer_Incidence_Rate.svg){#fig-pyramid-incidence}

![Disability-Adjusted Life Year](output/figures/Population pyramid - 2021/fig_C_United States of America_Bladder cancer_DALYs_Rate.svg){#fig-pyramid-dalys}

![Years Lived with Disability](output/figures/Population pyramid - 2021/fig_C_United States of America_Bladder cancer_YLDs_Rate.svg){#fig-pyramid-ylds}

![Years of Life Lost](output/figures/Population pyramid - 2021/fig_C_United States of America_Bladder cancer_YLLs_Rate.svg){#fig-pyramid-ylls}

![Mortality](output/figures/Population pyramid - 2021/fig_C_United States of America_Bladder cancer_Deaths_Rate.svg){#fig-pyramid-deaths}

Population pyramids visualizing age-gender distribution of 
bladder cancer burden rates in United States. 
:::

Similar population pyramids visualizing each state's 
burden rates across age groups in 2021
are available
from [Supporting Files](#sup-pyramid2021) and
the study's
[GitHub repository](https://github.com/alireza5969/bladder-cancer-gbd/tree/main/output/figures/Population%20pyramid%20-%202021).
In addition, population pyramid plots animated based on year of 
report are available for each each state and the country 
from [Supporting Files](#sup-pyramidgif) and
the study's
[GitHub repository](https://github.com/alireza5969/bladder-cancer-gbd/tree/main/output/figures/Population%20pyramid%20-%20animated).

@fig-apc, similar to @fig-pyramid, presents the 
burden rates for each age group. However,
it adds the element of time to the visualization. 
Similar plots are available for each state
from [Supporting Files](#sup-apc) and
the study's
[GitHub page](https://github.com/alireza5969/bladder-cancer-gbd/tree/main/output/figures/APC%20plot%201).


::: {#fig-apc layout-ncol=2}

![Prevalence](output/figures/APC plot 1/fig_F_United States of America_Prevalence_Bladder Cancer_Rate.svg){#fig-apc-prevalence}

![Incidence](output/figures/APC plot 1/fig_F_United States of America_Incidence_Bladder Cancer_Rate.svg){#fig-apc-incidence}

![Disability-Adjusted Life Year](output/figures/APC plot 1/fig_F_United States of America_DALYs_Bladder Cancer_Rate.svg){#fig-apc-dalys}

![Years Lived with Disability](output/figures/APC plot 1/fig_F_United States of America_YLDs_Bladder Cancer_Rate.svg){#fig-apc-ylds}

![Years of Life Lost](output/figures/APC plot 1/fig_F_United States of America_YLLs_Bladder Cancer_Rate.svg){#fig-apc-ylls}

![Mortality](output/figures/APC plot 1/fig_F_United States of America_Deaths_Bladder Cancer_Rate.svg){#fig-apc-deaths}

Age-gender distribution of 
bladder cancer burden rates across decades in United States. 
:::


## Bladder Cancer Burden Across the United States


```{r}


state_change_data <- 
  change_data %>% 
  filter(
    !is.na(stateabb),
    color != "neutral"
    ) %>% 
  mutate(
    sex_name = if_else(sex_id == 2, "women", "men"),
    across(diffval:difflower, \(x) return(abs(x) * 100))
    )
  
green_change_data <- 
  state_change_data %>% 
  filter(color == "negative")

red_change_data <- 
  state_change_data %>% 
  filter(color == "positive")

green_change_txt <- 
  glue_data(
    green_change_data %>%
      arrange(-diffval) %>%
      filter(location_name == "New York"),
    "{number(diffval, .1)}% [{number(difflower, .1)}%, {number(diffupper, .1)}%]"
    )

red_change_txt <- 
  glue_data(
    red_change_data,
    "{number(diffval, .1)}% [{number(difflower, .1)}%, {number(diffupper, .1)}%]"
    )

```

As shown in @tbl-percentchange, while some states such as New York
(YLLs: `r green_change_txt[2]` and `r green_change_txt[1]` 
in men and women, respectively; DALYs: `r green_change_txt[6]` 
and `r green_change_txt[4]` in men and women, respectively;
mortality: `r green_change_txt[5]` in women), 
with some of the highest change percentages among the states,
have managed to decrease BC burden statistics, most of the states 
show no significant change. The only state that has observed 
increased burden is New Mexico, which reports an increased
BC prevalence by `r red_change_txt` in men. 
State-specific, gender-stratified historical trends in 
ASIR are visualized in @fig-geofacet.
Similar graphs for age-standardized rates of
prevalence,
mortality, 
DALYs,
YLLs,
and YLDs
are available from [Supporting Files](#sup-geofacet) and
the study's
[GitHub repository](https://github.com/alireza5969/bladder-cancer-gbd/tree/main/output/figures/Geofacet%20plot).

![State-level historical trends in age-standardized prevalence rate across the states in men and women. States are positioned to nearly resemble their geographical location.](output/figures/Geofacet plot/fig_A_Incidence_Bladder cancer_Rate.svg){#fig-geofacet}

```
Abbreviations: `r paste0(c(state_abbs), collapse = "; ")`
```


# Discussion

This study used GBD 2021 data to analyze trends in BC incidence, mortality, YLLs, YLDs, and DALYs in
the United States from 1980 to 2021. Nationally, the ASIR of BC has gradually declined since the
late 2010s, accompanied by a less steep decrease in the ASMR. 
These improvements likely reflect
advancements in early detection strategies, enhanced treatment modalities, and broader access to
quality healthcare services across the nation. 
However, these favorable national trends mask
important state-level disparities. While many Northeastern states have experienced sharp declines in
both incidence and mortality, especially among men, certain states such as West Virginia,
Kentucky, Delaware, and Nevada have shown rising or plateauing ASIRs and ASMRs. 
Among the states with notable trends, some have persistently stood out in terms of exceptionally high or low rates.
Maine consistently exhibited the highest ASIR across the study period, with a declining trend in
recent years, whereas Hawaii showed the lowest ASIR and ASMR [Supporting Files](@sup-geofacet).
These differences may reflect regional
variations in exposure to risk factors, healthcare access, and public health interventions.

Despite having one of the highest per capita health expenditures globally [@RN24], the country also
reports one of the highest ASIR of BC compared to many high-income and high-SDI regions, a
finding that is consistent with recent research [@RN3]. This discrepancy likely reflects the United
States' advanced diagnostic infrastructure and widespread access to healthcare services [@RN25],
which facilitate superior case detection rather than a true increase in disease burden [@RN26;
@RN27]. However, this also underscores a potential misalignment in healthcare priorities, where
substantial resources may be disproportionately directed toward treatment and diagnostics rather
than prevention and risk factor reduction [@RN28]. The persistently high incidence rates call for
stronger public health strategies, given that recent studies indicate that approximately 80% of BC
cases diagnosed over the past two decades are likely associated with identifiable preventable
factors [@RN29]. In contrast, a study suggested that inherited genetic factors are estimated to
account for only 7% of BC incidences in western populations [@RN30]. Given the substantial
contribution of known environmental risk factors, BC represents a prime candidate for targeted
public health policy preventive efforts, particularly in smoking cessation, occupational hazard mitigation,
and environmental exposure reduction [@RN31].

The age distribution analysis of BC incidence, DALYs, and mortality across the states from 1990 to
2021 reveals a consistent pattern of increasing burden with advancing age, particularly
in older age groups, consistent with previous findings [@RN3]. While age-specific trends 
have remained relatively stable over time, 2021 data showed slightly elevated incidence and mortality in the
oldest age groups across several states, including California, Texas, and Connecticut [Supporting Files](@sup-apc). 
This can be partly due to the substantial aging of the overall population during this period, marked by a
38.6% increase in the number of people aged 65 and older from 2010 to 2020 [@RN31]. In addition,
recent trends show the oldest Americans (especially men 85+) are not experiencing the same mortality
improvements as younger seniors. Notably, the most significant rise occurred in the 75-79 and 80+
age groups, suggesting the cumulative impact of lifetime exposures and the biological role of aging
[@RN32]. These findings suggest that while early detection and treatment may be improving overall
outcomes, the burden in the elderly remains a critical public health concern [@RN19].

Consistent with the broader literature, our own analysis confirms that men experience bladder burden rates
of roughly four-fold higher than women [@RN22]. When we examined age-period curves
across multiple states, the shapes of the age‐specific burden trajectories were
somewhat similar for men and women in the sense that both sexes showed an exponential 
rise after age 60, despite
women having much lower rates and ASMR has fallen more slowly in women than men. In
addition, studies have shown that female patients often present with more advanced disease and
poorer outcomes [@RN33; @RN23; @RN21]. These gender disparities are not unexpected, and one
explanation is that they may partly stem from previous studies suggesting that women with hematuria
often experience delays in evaluation, as the symptom is frequently misattributed to urinary tract
infections or gynecologic conditions [@RN33; @RN34]. To address these disparities, clinicians should
implement the 2025 American Urological Association (AUA) guideline 
updates [@RN35] in routine practice and consider the use of
validated biomarkers for risk stratification, while primary care organizations are encouraged to
adopt sex‑aware diagnostic protocols and systematically monitor referral patterns for hematuria in
women.

In addition, emerging research highlights key biological contributors to these disparities: men and
women differ in immune surveillance, circulating hormone levels (and hormone receptor expression),
and genetic/epigenetic alterations [@RN36]. For example, male hormone signaling may promote certain
cancer pathways, while female biology (e.g. different immune responses) may influence tumor
aggression [@RN37; @RN38]. Importantly, these intrinsic differences persist even after adjusting for
known exposures, such as smoking and occupation [@RN8]. Ongoing studies of sex hormones, immune
checkpoints, and genomic profiles aim to explain why men are so much more likely to develop bladder
cancer and why women exhibit worse outcomes once diagnosed.

Regarding geographic and demographic disparities in BC burden, consistent with our study,
studies from United States
show the highest incidence among males and in the Northeast region (e.g. Maine, New
Hampshire, Vermont) [@RN16]. This pattern cannot be fully explained by smoking alone [@RN12]. In
fact, researchers have linked elevated incidence in these New England states to environmental
carcinogens, one of which is arsenic in drinking water using private domestic wells [@RN12; @RN13].
Declines in Northeast BC rates over the past decade suggest improvements in water treatment [@RN39]
and broader public health interventions. For example, in both Maine and New Jersey, household
arsenic treatment systems, particularly those professionally installed and maintained, have
achieved a substantial seven-fold reduction in arsenic exposure and associated cancer risk [@RN40].
However, nearly 20% of these systems failed to reduce arsenic to acceptable levels, underscoring the
importance of proper installation, routine maintenance, and ongoing monitoring to ensure sustained
effectiveness [@RN40]. Another example is in an effort to reduce arsenic-related health risks such
as BC, New Hampshire lowered its public drinking water arsenic standard to 5 µg/L, which is 
half the federal Environmental Protection Agency (EPA)
limit, through a law passed in 2019 and implemented in 2021 [@RN41]. Despite public health
efforts, these states continue to lead in BC incidence partly due to the disease's long latency and
lifetime cumulative arsenic exposure. Recent reductions in arsenic levels will take time to
translate into lower rates, especially as many private wells still contain legacy contamination.

By contrast, certain other regions exhibit disproportionately high mortality from BC even if
incidence is not highest. For example, we found that Nevada and some southern states like West
Virginia, Kentucky, and Alabama show some of the nation's highest ASMRs from BC, similar to previous
studies [@RN42; @RN19]. The elevated BC mortality observed in the Southern states may be partly
explained by the substantial burden of smoking-attributable cancer deaths in this region [@RN43]. In
addition, the higher proportion of rural areas in southern states may contribute to increased
mortality risk [@RN44; @RN45], potentially due to limited access to healthcare services and delayed
cancer detection or treatment [@RN45]. Limited availability of healthcare providers and the need for
long-distance travel to treatment centers can cause further delays [@RN46], while lower insurance
coverage rates among rural residents have been reported as contributing factors to diagnostic delays
in other cancers [@RN47]. On the contrary, Northeast's low uninsured rate [@RN48] may contribute to
relatively lower BC mortality, despite the region's relatively high incidence rates. As a result, a
regional "mismatch" becomes evident as states like those in New England exhibit high incidence but
relatively low mortality, while others in the South and West experience persistently high death
rates despite comparable or lower incidence.

Hawaii stands out with one of the lowest state-level BC burden rates and trends. Several factors likely
contribute to Hawaii's protective profile. The population is predominantly Asian/Pacific Islander,
groups that historically have lower BC risk than non-Hispanic Whites [@RN49]. Hawaii also boasts low
adult smoking prevalence [@RN50] and strong tobacco-control policies relative to many mainland
states [@RN51]. Environmental exposures such as industrial chemicals, arsenic, pollutants, are generally
less concentrated in Hawaii. In combination, these demographic and lifestyle differences along
with relatively good access to healthcare and high rates of health insurance [@RN52] likely
explain why Hawaii has the least BC burden despite excellent longevity. In summary, Hawaii's example
underscores how reduced exposure to known risks (e.g. tobacco and toxins) and access to care can produce
strikingly lower BC incidence and mortality.

Strengths of our study include in the use of robust population-based data to examine
state-level disparities and temporal trends in BC incidence and mortality across the United States.
This study has several limitations that warrant consideration. The GBD
analysis is constrained by the limited availability of high-quality primary data, necessitating the
use of modeling techniques that may compromise predictive validity in settings where data are sparse
or inconsistent. While the sophisticated data processing and statistical modeling methods employed
by GBD collaborators help mitigate this issue, they also introduce an inherent level of uncertainty
into the estimates. In addition, there is a lag in the availability of GBD data. As our analysis is
based on data spanning 1990 to 2021, it may not fully capture the most recent trends in the burden
of BC. Lastly, this study focuses primarily on the epidemiological patterns of BC and does not
incorporate analyses of specific clinical aspects and risk factors, thereby limiting
the ability to draw causal inferences.

# Conclusions

BC remains a significant public health challenge in the United States, marked by notable geographic,
demographic, and sex-based disparities. Despite national declines in some burden measures,
certain states continue to bear a disproportionate burden, driven by environmental exposures, aging
populations, and limited healthcare access. Our analysis highlights the need for region-specific
prevention policies, improved diagnostic equity, and robust surveillance systems. Addressing
modifiable risk factors alongside expanding access to early detection and care, is crucial. Targeted
public health initiatives and policies are essential to reduce disparities and alleviate the
long-term burden of this costly and preventable disease.

---

> **Comments by Fatemeh**
> 
> [Delaware, and Nevada have shown rising or plateauing ASIRs and ASMRs. The trends shown in CDC
except for NV ASMR are declining, probably because it's not age-adjusted. Anyway, focus on NV
ASMR]
> 
> [States like the District of Columbia exhibited higher incidence and DALY rates in middle-aged males
(50--69), potentially indicating early-life exposure to carcinogens or disparities in early
diagnosis. This is not age adjusted, it can be just an increase in population]
> 
> [Among northeastern states, New York shows to have a shaper decline in ASIR and ASMR. (this may not
be a good comparison, because of the rural population has more incidence and NYers don't use high
domestic wells probably)]
> 
> [Health Economics Research: Conduct health economic analyses to assess the cost-effectiveness of
shifting healthcare resources towards primary prevention and early detection, compared to the
current disproportionate investment in late-stage treatment. Such research would provide crucial
evidence for policy decisions aimed at optimizing public health spending and improving long-term
outcomes for BC patients nationwide.]

---


# Supporting Files {.unnumbered}

Supporting Files are available as appendices to the manuscript and 
from the following URLs.

1. Plots for historical trends (geographically faceted by states) 
are available from the project's
[GitHub repository](https://github.com/alireza5969/bladder-cancer-gbd/tree/main/output/figures/Geofacet%20plot).[](){#sup-geofacet}

2. Population pyramid plots of each state's age- and gender-stratified
burdens in 2021 are available from the project's
[GitHub repository](https://github.com/alireza5969/bladder-cancer-gbd/tree/main/output/figures/Population%20pyramid%20-%202021).[](){#sup-pyramid2021} 

3. Animated historical population pyramid plots of each state's
age- and gender-stratified
burdens are available from the project's
[GitHub repository](https://github.com/alireza5969/bladder-cancer-gbd/tree/main/output/figures/Population%20pyramid%20-%20animated).[](){#sup-pyramidgif} 

4. Bump plots ranking states based on their 
burden statistics in 1990, 2000, 2010, and 2021 are available 
from the project's 
[GitHub repository](https://github.com/alireza5969/bladder-cancer-gbd/tree/main/output/figures/Bump%20plot).[](){#sup-bump} 

5. Line plots visualizing each state's age- and gender-stratified burdens
across decades are available from the project's 
[GitHub repository](https://github.com/alireza5969/bladder-cancer-gbd/tree/main/output/figures/APC%20plot%201).[](){#sup-apc}


# Statements {.appendix}

## Contributions {.appendix}

```{r}
#| label: contrib
#| echo: false
#| message: false
#| warning: false
#| results: asis


authors <- rmarkdown::metadata$author

df <- 
  purrr::map_dfr(authors, function(a) {
    tibble::tibble(
      name = a$name,
      role = unlist(a$roles)
      )
    })

role_order <- df |>
  dplyr::distinct(role) |>
  dplyr::pull(role)

for (r in role_order) {
  people <- df |>
    dplyr::filter(role == r) |>
    dplyr::pull(name)

  role_label <- stringr::str_to_sentence(r)

  cat(
    "**", role_label, ":** ",
    paste(people, collapse = ", "),
    "; ",
    sep = ""
  )
}
```

## Acknowledgments {.appendix}

The authors declare no conflicts of interest.

## Transparency Statement {.appendix}

The lead author, Alireza Sadeghi, affirms that this manuscript is an honest, accurate, and
transparent account of the study being reported; that no important aspects of the study have been
omitted; and that any discrepancies from the study as planned (and, if relevant, registered) have
been explained.

## Data Availability Statement {.appendix}

The data utilized in this study are publicly available through the [Global Burden of Disease (GBD) Study 2021 repository](http://ghdx.healthdata.org/gbd-results-tool). 
The data and code used to produce this manuscript 
are available from the study's 
[GitHub repository](https://github.com/alireza5969/bladder-cancer-gbd). 
Data is available in R's native binary file format (RDS).
To reproduce the manuscript and the results, 
clone the project repository and render *index.qmd* file.

## Conflicts of Interest {.appendix}

The authors declare no conflicts of interest.

# References {.unnumbered}

::: {#refs}
:::

:::::